---
title: "MA563 HW4"
author: "Maggie Zhang (cz239)"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4)
```

## Problem 1: Discrete Least Squares

Let $x_j = jh$ with $h = \frac{2\pi}{N}$ (for $j = 0, \ldots, N-1$) be
the usual nodes for the DFT.

------------------------------------------------------------------------

### 1. Verify the Discrete Orthogonality Relation

We need to show that $\langle e^{ijx}, e^{ikx} \rangle_d = 0$ for
$j \neq k$, where
$\langle f, g \rangle_d = \sum_{j=0}^{N-1} f(x_j) \overline{g(x_j)}.$

**Sol:**

Compute:
$$\langle e^{ijx}, e^{ikx} \rangle_d = \sum_{\ell=0}^{N-1} e^{ijx_\ell} \, \overline{e^{ikx_\ell}} = \sum_{\ell=0}^{N-1} e^{ij\ell h} e^{-ik\ell h} = \sum_{\ell=0}^{N-1} e^{i(j-k)\ell h}.$$

Let $d = j-k$. Then the sum is a geometric series:
$$S = \sum_{\ell=0}^{N-1} \left(e^{idh}\right)^\ell.$$

Since $h = 2\pi/N$, we have $e^{idhN} = e^{i d 2\pi} = 1$.

-   If $d$ is a multiple of $N$, then $e^{idh}=1$ and each term equals
    1, so $S = N$.
-   Otherwise, $e^{idh} \neq 1$ and the geometric series formula gives
    $$S = \frac{1 - e^{idhN}}{1 - e^{idh}} = \frac{1-1}{1 - e^{idh}} = 0.$$

Thus for $j \neq k$ with $|j-k|$ not a multiple of $N$, the inner
product is zero. In particular, for frequencies in the range $[-m,m]$
with $m < N/2$, this condition holds, so the exponentials are orthogonal
on the discrete set $\{x_j\}$.

------------------------------------------------------------------------

### 2. Least Squares Approximation

We are given values $f_j = f(x_j)$ at the nodes and seek an
approximation
$$p_m(x) = \sum_{k=-m}^{m} c_k e^{ikx}, \quad m < \frac{N}{2},$$ that
minimizes the discrete least squares error
$$E(\vec{c}) = \sum_{j=0}^{N-1} \bigl|p_m(x_j) - f_j\bigr|^2.$$

Interpret the data as a vector
$\mathbf{f} = (f_0,\dots,f_{N-1})^T \in \mathbb{C}^N$ and the basis
functions as vectors
$$\boldsymbol{\phi}_k = \bigl(e^{ikx_0}, e^{ikx_1}, \dots, e^{ikx_{N-1}}\bigr)^T, \quad k = -m,\dots,m.$$\
From part 1, these vectors are orthogonal with respect to the standard
Hermitian inner product on $\mathbb{C}^N$:
$$\langle \boldsymbol{\phi}_k, \boldsymbol{\phi}_\ell \rangle = \sum_{j=0}^{N-1} e^{ikx_j} \overline{e^{i\ell x_j}} = 0 \quad (k \neq \ell),$$\
and their squared norms are\
$$\|\boldsymbol{\phi}_k\|^2 = \sum_{j=0}^{N-1} 1 = N.$$

The error $E$ is the squared distance between $\mathbf{f}$ and the
linear combination $\sum_{k} c_k \boldsymbol{\phi}_k$. In a
finite-dimensional inner product space, the best approximation
(minimizing the distance) is given by the orthogonal projection onto the
subspace spanned by $\{\boldsymbol{\phi}_k\}$. The coefficients satisfy
the normal equations
$$\sum_{\ell=-m}^{m} c_\ell \langle \boldsymbol{\phi}_\ell, \boldsymbol{\phi}_k \rangle = \langle \mathbf{f}, \boldsymbol{\phi}_k \rangle \quad \text{for each } k.$$
Due to orthogonality, this reduces to
$$c_k \langle \boldsymbol{\phi}_k, \boldsymbol{\phi}_k \rangle = \langle \mathbf{f}, \boldsymbol{\phi}_k \rangle,$$
so
$$c_k = \frac{\langle \mathbf{f}, \boldsymbol{\phi}_k \rangle}{\|\boldsymbol{\phi}_k\|^2} = \frac{1}{N} \sum_{j=0}^{N-1} f_j \, \overline{e^{ikx_j}} = \frac{1}{N} \sum_{j=0}^{N-1} f_j \, e^{-ikx_j}.$$
This is precisely the (scaled) discrete Fourier transform coefficient
for frequency $k$. Thus the least squares solution yields the DFT
coefficients.

------------------------------------------------------------------------

## Problem 2: Testing of Discrete Fourier Transform

### (a) Fourier Series Approximation

We write a function that computes the partial sum $S_N$ of the Fourier
series for a given function $f$ on $[0, 2\pi]$, then apply it to
$f(x) = x$ and $f(x) = \sin(2x)$, plot $S_N(x)$, and plot the $L^2$
error $\|S_N - f\|_{L^2}$ versus $N$.

**Fourier series used:**

\- For $f(x) = x$ on $[0, 2\pi]$: $a_0/2 = \pi$, $b_n = -\frac{2}{n}$
for $n \ge 1$, so $S_N(x) = \pi - \sum_{n=1}^{N} \frac{2}{n} \sin(nx)$.

\- For $f(x) = \sin(2x)$: the Fourier series has only the term
$\sin(2x)$, so $S_N(x) = \sin(2x)$ for $N \ge 2$.

```{r 2a-fourier-partial-sum}
# Partial sum S_N of Fourier series for f(x) = x on [0, 2*pi]
# S_N(x) = pi - sum_{n=1}^N (2/n) sin(n x)
S_N_x <- function(x, N) {
  if (N < 1) return(rep(pi, length(x)))
  s <- pi
  for (n in 1:N) s <- s - (2 / n) * sin(n * x)
  s
}

# Partial sum for f(x) = sin(2x): just sin(2x) for N >= 2
S_N_sin2x <- function(x, N) {
  if (N < 2) {
    # N=0: zero, N=1: only sin(x) coefficient would be 0 for sin(2x)
    return(rep(0, length(x)))
  }
  sin(2 * x)
}

# L2 norm of (g - f) on [0, 2*pi]: sqrt(integral_0^{2pi} |g(x)-f(x)|^2 dx)
L2_error <- function(g_fun, f_fun, N_vals, x_fun = S_N_x, n_points = 1000) {
  x <- seq(0, 2*pi, length.out = n_points)
  dx <- (2*pi) / (n_points - 1)
  err <- numeric(length(N_vals))
  for (i in seq_along(N_vals)) {
    N <- N_vals[i]
    g <- x_fun(x, N)
    f <- f_fun(x)
    err[i] <- sqrt(sum(abs(g - f)^2) * dx)
  }
  err
}
```

```{r 2a-plot-fx-equals-x}
# f(x) = x 
x_plot <- seq(0, 2*pi, length.out = 500)
N_vals_plot <- c(1, 3, 5, 10, 20)

# Graph 1: S_N(x) vs f(x) = x
plot(x_plot, x_plot, type = "l", lwd = 2, col = "black",
     xlab = "x", ylab = "y", main = expression("f(x)=x:"~S[N](x)~"vs f(x)"))

# Adding the Fourier series partial sums
for (i in 1:length(N_vals_plot)) {
  N <- N_vals_plot[i]
  lines(x_plot, S_N_x(x_plot, N), col = i + 1, lty = 2)
}

legend("bottomright", legend = c("f(x)", paste0("N=", N_vals_plot)),
       col = c("black", 2:(length(N_vals_plot)+1)), 
       lty = c(1, rep(2, length(N_vals_plot))),
       cex = 0.8)

# Graph 2: L2 error vs N for f(x)=x 
N_seq <- 1:50
err_x <- L2_error(S_N_x, function(x) x, N_seq, x_fun = S_N_x)

plot(N_seq, err_x, type = "o", pch = 16, cex = 0.6,
     xlab = "N", ylab = expression(group("|", S[N]-f, "|")[L^2]),
     main = expression("L2 error vs N (f(x)=x)"))

```

```{r 2a-plot-sin2x}
# f(x) = sin(2x)

# Graph 1: S_N(x) vs f(x) = sin(2x) 
plot(x_plot, sin(2*x_plot), type = "l", lwd = 2, col = "black",
     xlab = "x", ylab = "y", main = expression("f(x)=sin(2x):"~S[N](x)~"vs f(x)"))

# Adding the Fourier series partial sums
for (N in c(2, 5, 10)) {
  lines(x_plot, S_N_sin2x(x_plot, N), col = N, lty = 2)
}

legend("topright", legend = c("f(x)", "S_N (N>=2)"), 
       col = c("black", 2), lty = c(1, 2), cex = 0.8)

# Graph 2: L2 error vs N for f(x)=sin(2x)
N_seq2 <- 1:20
err_sin2x <- L2_error(S_N_sin2x, function(x) sin(2*x), N_seq2, x_fun = S_N_sin2x)

plot(N_seq2, err_sin2x, type = "o", pch = 16, cex = 0.6,
     xlab = "N", ylab = expression(group("|", S[N]-f, "|")[L^2]),
     main = expression("L2 error vs N (f(x)=sin(2x))"))
```

**Observations (a):**

\- **f(x) = x:** Gibbs Phenomenon: As $N$ increases, the partial sums
$S_N(x)$ track the linear slope of $f(x)$ more closely in the center of
the interval. However, because the periodic extension of $f(x) = x$ has
a jump discontinuity at the boundaries ($0$ and $2\pi$), prominent
oscillations persist near these edges. The plot of the $L^2$ error shows
a steady, gradual decrease as $N$ increases. The error does not drop to
zero quickly because the higher-frequency components are necessary to
attempt to resolve the sharp discontinuity at the boundary.

\- **f(x) = sin(2x):** For $N \ge 2$, the approximation $S_N(x)$ is
visually indistinguishable from the true function $f(x)$. Since
$\sin(2x)$ is itself one of the basis functions used in the Fourier
series, the approximation becomes exact as soon as that frequency is
included in the sum. The $L^2$ error plot confirms this "perfect fit".
The error is high for $N=1$ , but it drops immediately to zero for $N=2$
and remains at zero for all subsequent values of $N$. This demonstrates
that finite Fourier series can perfectly reconstruct functions composed
of a finite number of sinusoids.

------------------------------------------------------------------------

### (b) DFT Implementation (Without FFT)

We compute the DFT coefficients
$F_k = \frac{1}{N} \sum_{j=0}^{N-1} f(x_j) e^{-ik x_j}$ with
$x_j = \frac{2\pi j}{N}$, without using built-in FFT. We choose a
function, compute $F_k$, and plot real and imaginary parts to observe
aliasing.

```{r 2b-dft-by-hand}
# DFT: F_k = (1/N) sum_{j=0}^{N-1} f(x_j) exp(-1i * k * x_j), x_j = 2*pi*j/N
dft_manual <- function(f_vals) {
  N <- length(f_vals)
  k_vals <- 0:(N - 1)
  F_k <- complex(length(k_vals))
  for (k in k_vals) {
    F_k[k + 1] <- 0
    for (j in 0:(N - 1)) {
      x_j <- 2 * pi * j / N
      F_k[k + 1] <- F_k[k + 1] + f_vals[j + 1] * exp(-1i * k * x_j)
    }
    F_k[k + 1] <- F_k[k + 1] / N
  }
  F_k
}

# Example: f(x) = sin(2x) sampled at N points. For N small, aliasing occurs.
# Another example: f(x) = cos(5x) with N = 8 (few samples) to see aliasing
N_small <- 8
x_nodes <- (0:(N_small - 1)) * (2 * pi / N_small)
f_cos5 <- cos(5 * x_nodes)  # frequency 5 with only 8 points
F_cos5 <- dft_manual(f_cos5)

k_plot <- 0:(N_small - 1)
plot(k_plot, Re(F_cos5), type = "h", lwd = 2, col = "darkblue",
     xlab = "k", ylab = "Re(F_k)", main = "Real part of F_k (f(x)=cos(5x), N=8)")
points(k_plot, Re(F_cos5), pch = 16, col = "darkblue")

plot(k_plot, Im(F_cos5), type = "h", lwd = 2, col = "darkred",
     xlab = "k", ylab = "Im(F_k)", main = "Imaginary part of F_k (f(x)=cos(5x), N=8)")
points(k_plot, Im(F_cos5), pch = 16, col = "darkred")
```

**Aliasing explanation (b):** We use $f(x) = \cos(5x)$ with only $N = 8$
samples. The true frequency is 5, but with 8 samples the DFT only has
indices $k = 0, \ldots, 7$. For a real signal,
$F_{N-k} = \overline{F_k}$, so energy at high frequency “folds back.”
Here $\cos(5x) = \frac{1}{2}(e^{i5x}+e^{-i5x})$; with period $2\pi$,
mode 5 is equivalent to mode $5 - 8 = -3$, so we see nonzero
coefficients at $k = 3$ and $k = 5$ (and by symmetry at $k = 8-3=5$). So
the high frequency (5) is aliased to a lower index. This illustrates
that if we sample below the Nyquist rate (here we need at least 10
points to resolve frequency 5), the DFT coefficients do not uniquely
represent the original continuous frequency.

------------------------------------------------------------------------

### (c) Trigonometric Interpolation Using DFT Coefficients

Using the $F_k$ from part (b) for $f(x) = \sin(2x)$, we form the
trigonometric interpolant $p_N(x) = \sum_{k=0}^{N-1} F_k e^{ikx}$ and
verify that it interpolates the data $(x_j, f(x_j))$, and plot the
interpolation error $|p_N(x) - f(x)|$.

```{r 2c-trig-interpolant}
# For f(x) = sin(2x): sample at x_j = 2*pi*j/N, compute F_k, then p_N(x) = sum_k F_k exp(i k x)
N <- 16
x_j <- (0:(N - 1)) * (2 * pi / N)
f_vals <- sin(2 * x_j)
F_k <- dft_manual(f_vals)

# Trigonometric interpolant: p_N(x) = sum_{k=0}^{N-1} F_k exp(i k x)
p_N <- function(x, F_k) {
  N <- length(F_k)
  out <- 0 + 0i
  for (k in 0:(N - 1)) {
    out <- out + F_k[k + 1] * exp(1i * k * x)
  }
  Re(out)  # for real data, imaginary part is numerical noise
}

# Verify at nodes: p_N(x_j) should equal f(x_j)
p_at_nodes <- numeric(N)
for (j in 1:N) p_at_nodes[j] <- p_N(x_j[j], F_k)
cat("Max interpolation error at nodes:", max(abs(p_at_nodes - f_vals)), "\n")

# Plot interpolation error over [0, 2*pi]
x_fine <- seq(0, 2*pi, length.out = 500)
f_true <- sin(2 * x_fine)
p_fine <- sapply(x_fine, function(x) p_N(x, F_k))
err_interp <- abs(p_fine - f_true)

plot(x_fine, p_fine, type = "l", col = "blue", lwd = 2,
     xlab = "x", ylab = "y", main = "p_N(x) vs f(x)=sin(2x)")
lines(x_fine, f_true, col = "gray", lty = 2, lwd = 1)
points(x_j, f_vals, pch = 16, col = "red")
legend("topright", legend = c("p_N(x)", "f(x)", "data (x_j, f(x_j))"),
       col = c("blue", "gray", "red"), lty = c(1, 2, NA), pch = c(NA, NA, 16))

plot(x_fine, err_interp, type = "l", col = "purple", lwd = 1.5,
     xlab = "x", ylab = "Error(x)", main = "Interpolation error |p_N(x) - f(x)|")

```

**Interpretation (c):** The trigonometric interpolant $p_N(x)$ achieves
perfect accuracy at the discrete nodes, with a maximum error of
$4.88 \times 10^{-15}$ reflecting the limits of machine precision. While
$p_N(x)$ is a smooth $2\pi$-periodic polynomial that correctly
interpolates the data points, significant error occurs between nodes
because the summation uses only high positive frequencies ($k=0$ to
$N-1$) rather than centering them around zero. In this case, with
$N=16$, the error $\|p_N(x) - f(x)\|$ remains zero at the nodes but
oscillates in between due to the lack of negative frequency components
necessary to reconstruct the smooth $\sin(2x)$ signal. This demonstrates
that while the DFT ensures exactness at sampled points, the
representation of the function between those points depends heavily on
whether the sampling rate is sufficient to resolve the underlying
frequencies without aliasing.
